---
title: "Sílabas"
author: "Juliana Zapata"
date: "1/10/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Manejo de datos, gráficos, colores
library(readxl)
library(tidyverse)
library(jcolors)
library(DT)
library(plotly)
library(ggpubr)
library(broom)
library(janitor) # edición de nombres
library(mgcv) # Modelos GAM
library(psych)
library(autoplotly)


# Análisis multivariado
library(FactoMineR)
library(factoextra)
```

```{r}
datos<- read_excel("~/Desktop/Estadistica Naty/BD tipo de silabas.xlsx", 
    sheet = "Total")

datos
```
# Descriptivo y Exploratorio {.tabset .tabset-fade .tabset-pills}

## Descriptivo por grupo

```{r}
datos %>% 
  gather(key = "variable", value = "valor", -Cepa) %>% 
  group_by(variable, Cepa)  %>%
  summarise(Mínimo = round(min(valor, na.rm = TRUE), digits = 2),
            Media  = round(mean(valor, na.rm = TRUE), digits = 2),
            Máximo = round(max(valor, na.rm = TRUE), digits = 2),
            DE     = round(sd(valor, na.rm = TRUE), digits = 2),
            Q1     = round(quantile(valor, probs = 0.25, na.rm = TRUE), digits = 2),
            Q2     = round(quantile(valor, probs = 0.50, na.rm = TRUE), digits = 2),
            Q3     = round(quantile(valor, probs = 0.75, na.rm = TRUE), digits = 2),
            n = n()) %>% 
  ungroup() %>% 
  datatable(extensions = 'Buttons', options = list(
     dom = 'Bfrtip',
     buttons = c('csv', 'excel', 'pdf')))
```

## Distribuciones

```{r,fig.width=9, fig.height=4}
datos%>%
  gather(key = "variable", value = "value", -Cepa) %>% 
  
  mutate (variable = factor(variable, levels = c( "Armónicos	Ascendente", 	"Chevron descendente",	
                                    "Chevron",	"Corta",	"Compleja",	"Compleja con descendente",
                                    "Compleja ascendente",	"Descendente",	"Forma de U",
                                    "Forma de U invertida",	"Un salto",	"Un salto (largo)",	
                                    "Saltos múltiples",	"Saltos múltiples largos",	"Plana",	
                                    "Onda"))) %>% 
  group_by(variable, Cepa) %>% 
  summarise(promedio = mean(value, na.rm = TRUE),
            de = sd(value, na.rm = TRUE))  %>% 
  
  ggplot(aes(x = Cepa, y = promedio, color = Cepa,
             ymin  = promedio - de,
             ymax = promedio + de)) +
facet_wrap(~variable, scales = "free", ncol = 4) + 
  geom_point() +
  geom_errorbar(width = 0.2) +
  scale_fill_jcolors(palette = "pal7") +
  theme_light() +
  theme(strip.background = element_rect(fill = "snow4")
   )
```
# Gráfico 1

```{r, fig.width=5, fig.height=3}
library(jcolors)
datos %>% 
  gather(key = "variable", value = "value", -Cepa) %>% 
  
  mutate (variable = factor(variable, levels = c( "Armónicos	Ascendente", 	"Chevron descendente",	
                                    "Chevron",	"Corta",	"Compleja",	"Compleja con descendente",
                                    "Compleja ascendente",	"Descendente",	"Forma de U",
                                    "Forma de U invertida",	"Un salto",	"Un salto (largo)",	
                                    "Saltos múltiples",	"Saltos múltiples largos",	"Plana",	
                                    "Onda"))) %>% 
  group_by(Cepa, variable) %>% 
  summarise(promedio = mean(value, na.rm = TRUE),
            de = sd(value, na.rm = TRUE))  %>% 
  ggplot(aes(x = Cepa, y = promedio, color = Cepa)) +
  scale_color_manual(values = c("orange", "skyblue")) +
  facet_wrap(~variable, scales = "free") +
  geom_point(size = 2, colour="grey50") +
  theme_classic()+
  geom_line(size = 0.6, linetype =1) +
  theme(legend.position = "top") +
  labs(x = "Week of age", y = "Frequency of skin lesions ", color = "")
```
# Distribuciones originales

```{r, fig.width=9, fig.height=8}
datos %>% 
  select_if(is.numeric) %>% 
  gather(key = "variable", value = "valor") %>% 
  ggplot(data = ., aes(x = valor)) +
  facet_wrap(~variable, scales = "free") +
  geom_density()
```


```{r}
datos
datos$Cepa <- factor (datos$Cepa )
```
#Test z
```{r}
prop.test(x = c(12,14), n = c(225, 261),alternative = "two.sided")
```
Two-tailed Test porque Null Hypothesis: Both population proportions are equal.H0 : P1 = P2 y Alternate Hypothesis: Both population proportions are not equal.Ha: P1 ≠ P2. Level of significance (α) = 0.05
p-value: This is the p-value corresponding to a statistic.
p-value: This is the p-value corresponding to a statistic. Example with p-value is 0.003913.In this case, an alternative hypothesis is population proportions are not equal, i.e. two-tailed.
95 percent confidence interval: This gives us a 95% confidence interval for the true proportion. Here the 95% confidence interval is [0.0152883,0.0847117].
sample estimates: It gives the sample proportions. In our case, prop 1 = 0.21 and prop 2 = 0.16.
The rejection criteria for the given confidence level and conclude the results whether the test statistic lies in the rejection region or non-rejection region.

Conclusion:

Since the p-value[0.003913] is less than the level of significance (α) = 0.05, we reject the null hypothesis.

This means we have sufficient evidence to say that the population proportions are not equal.







